Noisy Work: A Review of The Uncertainty Visualisation Literature

### 2024-02-02 paper.qmd
made some edits to the general attitudes

# 3.General Attitude Towards Uncertainty (Uncertainty Communication Boradly)
It would be impossible to discuss uncertainty visualisation without mentioning the concept of uncertainty communication and discussing how visualisation fits into that broader scope. Uncertainty visualisation is related to uncertainty communication in the same way that visualisation more broadly is related communication more broadly. Visualisation is fast, it can communicate more complicated ideas that what can be done with text alone, and

*psychology of uncertainty communication*

It is difficult to read a paper that discusses uncertainty that does not make a nod towards uncertainty bias. In the case of uncertainty visualisation, these nods are not always relevant. Similarly to how contextual information is conflated with uncertainty when we try to define it, human attitudes to risk are often commented on in the discussions of uncertainty visualisation. Ironically, despite being mentioned in a large proportion of papers that discuss uncertainty risk aversion, has very to do with the psychology of uncertainty. As a matter of fact, it has so little to do with uncertainty communication that we would not even mention it in this paper were it not for the fact that ignoring it would suggest we had missed an important section of the literature.

Risk-aversion is the tendancy of people to prefer outcomes of low uncertainty to those outcomes with high uncertainty despute the outcomes with high uncertainty having a higher or equal expected outcome.

Uncertainty information is typically included in some form of statistical communication to offer some transparency to the viewer. Data analysis is a long and complicated process, so much so that even experts in the field can find themselves lost with a single incorrect assumption, so uncertainty can help pull back the curtain on this process and prevent people from understanding it as a human agnostic, entirely deterministic process. When science authors discuss "risk avoidance" in the uncertainty communication conversations, it holds with it some implication that heuristics such as risk avoidance are relevant in every conversation about uncertainty communication, however that is not true.

Risk and uncertainty are not perfect synonyms of one another. @Spiegelhalter2017 makes the distinction clear by specifying *risk* as *a precise random event* (such as a coin flip) while uncertainty is imprecision (such as the estimated probability associated with a coin that might be biased). Additionally, risk avoidance is observed when people have a choice between a *deterministic* outcome and a *risky* event that has higher expected value. How this psychological heuristic is therefore related to all uncertainty visualisation despite uncertainty communication being different in both the scenario (i.e. it is highly unlikely that we need to communicate a choice between a perfectly deterministic and an uncertain event) and in the type of uncertainty (since risk and uncertainty are not synonyms).

@@

- compares a bunch of stuff to the chance of getting struck by lightning since they are on a common scale (which is kind of funny) @Spiegelhalter2017
- hard to communicate small risks because people struggle to discrimintate between 1 in 10 000 and 1 in 100 000. Common solution is to use a logarithmic scale (if drawn vertically these are known as risk ladders). design of scale important because risk is associated with position rather than magnitude @Spiegelhalter2017
- using words to communicate numerical risk is generally discouraged and using words without numerical interpretation can lead to lower accuracy and higher risk aversion @Spiegelhalter2017
- the Affect heuristic comes to play in how people understand risk @Spiegelhalter2017
- avoids the topic of risk at times of accute crisis because numerical risk would be inappropriate @Spiegelhalter2017
- positive vs negative framing matters, can try to avoid it @Spiegelhalter2017
- no preference between percent or frequency although (small interaction with numeracy skills though) @Spiegelhalter2017
- Need careful specification of the reference class (e.g. % chance of rain should be fractions of situations like this in which is rains - over both time and location) @Spiegelhalter2017
- Need to make time interval clear (50% each year) @Spiegelhalter2017
- understanding better in a table than text @Spiegelhalter2017
- if expressing as a frequency, note that larger numerators suggest a larger risk, exteme version of this is ignoring the denominator all together @Spiegelhalter2017

- low-probability high-impact events are particularly difficult to communicate risks about @Spiegelhalter2017

# 4 LITERATURE SUMMARY
- methods used in uncertainty visualisation evaluation remain ad hoc. studies approach issue from a usability perspective instead of asking WHY representations DO or DO NOT work. Studies do not follow any methodology commonly agreed upon. @Kinkeldey2014
- Lack of formalisation and eigour in empirical methdos is an issue that is much broader than uncertainty visualisation (extends to info-vis and related domains). Therefore this paper is a starting point, not a definitive summary @Kinkeldey2014
- The question in these papers is often "does method a work better than method B" which is an engineering approach, where the toal is to improve a particular product rather than create general principals @Kinkeldey2014
- Authors then try to generalise beyond the specific constraints of the test even though the framing does not allow for it @Kinkeldey2014
- Alternatively, we should ask how/why does method A work better than method B, which grounds work in perceptual and cognative theory and provides a framwork @Kinkeldey2014

@@

# 4.?? Tasks and Motivating Questions (uncertainty about.. .what?)
- "cannot assess the quality of risk communication unless the objectives are clear" circles back to motivation. He assumes we are fulfilling a duty to inform. @Spiegelhalter2017
## 4.?? What are you uncertain about? A mathematical considerations
- Should add plot comparison mathematical framework example here
- take a plot and work through the entire example as a flow chart thing
## 4.1?? In other statistics
What is meant by "uncertainty" may seem obvious to some, but when you attempt to quantify or visualise it you will quickly find yourself asking, "uncertainty about... what?". Do you mean uncertainty on an estimate? On a forecast? How many steps ahead is this forecast? Are we only considering the uncertainty in the estimate or in the parameters or are we considering the possibility of measurement error or biased inputs? Signal and noise can only be untangled in the presence of a motivating question.

>>>>

# 3.General Attitude Towards Uncertainty (Uncertainty Communication Boradly)
It would be impossible to discuss uncertainty visualisation without mentioning the broader conversation around uncertainty communication. The conversation around uncertainty communication is often centered around the psychology is communicating risks.

Most research into general uncertainty communication seems to focus specifically on how we communicate risks. Risk and uncertainty are not perfect synonyms of one another. @Spiegelhalter2017 makes the distinction clear by specifying *risk* as *a precise random event* (such as a coin flip) while uncertainty is imprecision (such as the estimated probability associated with a coin that might be biased). For this reason, risk is often seen as a specific case of uncertainty and that well defined sub area has a lot of detailled research. Some markers of probability that have common uses, probabilities such as 0 (impossible), 0.5 (a fair coin flip), 1 (certain) are easy for people to have an internal sense of [@Spiegelhalter2017]. It is incredibly hard to communicate small risks, because people cannot differentiate between 1/100 an 1/1000, and communicating low probability but high impact risks are particularly challenging [@Spiegelhalter2017]. This issue can be somewhat alleviated by communicating the likelihood of an event in relation to another event on a similar scale (such as being struck by lightning) [@Spiegelhalter2017]. The framing of how we communicate risks matters due to the affect heuristic, so there will be a difference in peoples behaviours if, for example, you tell people they have a 90% chance of survival vs a 10% chance of death, and when expressing a frequency, a larger numerator communicates a larger risk [@Spiegelhalter2017]. The reference class of a particular risk should be explicitly stated to avoid confusion, for example if we are communicating the chance of rain, the time period over which we are expecting this chance of rain (i.e. 50% chance today or this week) needs to be specified [@Spiegelhalter2017]. The problem with a risk focus is that it is only a small piece of the uncertainty pie, and very few of these findings can be reasonably assumed to go beyond risk communication. Even less can be assumed to translate to uncertainty visualisation.

*this is all currently from the spieglehatter paper because that is the only paper I have moved notes over from mendeley before I started doing the summary write up. this will not all be from one paper, dont worry.*

Risk and uncertainty are quite different in practice. Uncertainty information is typically included in some form of statistical communication to offer some transparency to the viewer. Data analysis is a long and complicated process, so much so that even experts in the field can find themselves lost with a single incorrect assumption, so uncertainty can help pull back the curtain on this process and prevent people from understanding it as a human agnostic, entirely deterministic process. When science authors discuss "risk avoidance" in the uncertainty communication conversations, it holds with it some implication that heuristics such as risk avoidance are relevant in every conversation about uncertainty communication, however that is not true.

There are some elements of risk communication that seem to translate to uncertaitny communication more broadly. Using words to communicate risk (as well as uncertainty more broadly) is discouraged because people can misinterpret the actual risk involved, so when possible numerical estimates of risk should be communicated, however this can make information harder to understand for those with poor numeracy skills [@Spiegelhalter2017].

likely due uncertainty and risk being near synonymous, it is difficult to read a paper that discusses uncertainty that does not make a nod towards human attitudes towards risk. Even when the uncertainty being discussed in that particular paper has very little to do with risk.

Risk-aversion is the tendancy of people to prefer outcomes of low uncertainty to those outcomes with high uncertainty despute the outcomes with high uncertainty having a higher or equal expected outcome. The story of uncertainty communication is littered with conflated ideas, and that issue is highly visible here. Despite being mentioned in a large proportion of papers that discuss uncertainty risk aversion, seems to have very little to do communicating uncertainty. As a matter of fact, it has so little to do with uncertainty communication that we would not even mention it in this paper were it not for the fact that ignoring it would suggest we had missed an important section of the literature.

 Additionally, risk avoidance is observed when people have a choice between a *deterministic* outcome and a *risky* event that has higher expected value. How this psychological heuristic is therefore related to all uncertainty visualisation despite uncertainty communication being different in both the scenario (i.e. it is highly unlikely that we need to communicate a choice between a perfectly deterministic and an uncertain event) and in the type of uncertainty (since risk and uncertainty are not synonyms).

@@

Uncertainty visualisation is related to uncertainty communication in the same way that visualisation more broadly is related communication more broadly. Visualisation is fast, it can communicate more complicated ideas that what can be done with text alone, and 
# 4 LITERATURE SUMMARY
- methods used in uncertainty visualisation evaluation remain ad hoc. studies approach issue from a usability perspective instead of asking WHY representations DO or DO NOT work. Studies do not follow any methodology commonly agreed upon. @Kinkeldey2014
- Lack of formalisation and eigour in empirical methdos is an issue that is much broader than uncertainty visualisation (extends to info-vis and related domains). Therefore this paper is a starting point, not a definitive summary @Kinkeldey2014
- The question in these papers is often "does method a work better than method B" which is an engineering approach, where the toal is to improve a particular product rather than create general principals @Kinkeldey2014
- Authors then try to generalise beyond the specific constraints of the test even though the framing does not allow for it @Kinkeldey2014
- Alternatively, we should ask how/why does method A work better than method B, which grounds work in perceptual and cognative theory and provides a framwork @Kinkeldey2014

@@

# 4.?? Tasks and Motivating Questions (uncertainty about.. .what?)
- "cannot assess the quality of risk communication unless the objectives are clear" circles back to motivation. He assumes we are fulfilling a duty to inform. @Spiegelhalter2017
## 4.?? What are you uncertain about? A mathematical considerations
- Should add plot comparison mathematical framework example here
- take a plot and work through the entire example as a flow chart thing
- tasks e.g. "What is the average of X" (how well lined up you are moving down the data pipeline) and resolution (uncertainty is very often moving back a resolution level in statisitcs - also can be an assumption - bootstrap is a replacement for an assumption, not a resolution change, but bootstrapping is centered on the data)  e.g. "What is the noise around the estimate of the value of X"
## 4.1?? In other statistics
What is meant by "uncertainty" may seem obvious to some, but when you attempt to quantify or visualise it you will quickly find yourself asking, "uncertainty about... what?". Do you mean uncertainty on an estimate? On a forecast? How many steps ahead is this forecast? Are we only considering the uncertainty in the estimate or in the parameters or are we considering the possibility of measurement error or biased inputs? Signal and noise can only be untangled in the presence of a motivating question.
